\bf Title here

### 2020-07-20 paper/paper.tex
move notation to the first intro of PP; modify notation in the 3 algos

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. Let \(\mathbf{X}_{n \times p}\) be a
data matrix, an n-d projection can be seen as a linear transformation:
\(T: \mathbb{R}^p \mapsto \mathbb{R}^d\) defined by
\(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) be an index function
that maps the projection basis \(\mathbf{A}\) onto a scalar value \(I\),
this function is commonly known as the projection pursuit index (PPI)
function, or index function and can be used to measure the
``interestingness'' of the projection. A number of indice functions have
been proposed in the literature and this includes Legendre index
\citep{friedman1974projection}, hermite index
\citep{hall1989polynomial}, natural hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}. {[}Any literature on holes index?
xxx{]}

@@

Projection pursuit is usually used in conjunction with a tour method
called \emph{guided tour}. Tour explores the multivariate data
\emph{interactively} via playing a series of projections, that form a
\emph{tour path} and guided tour uses the path that is geodesically the
shortest. Details of the mathematical construction of a tour path can be
found in \citet{buja2005computational}. Modified from
\citet{buja2005computational}, Figure \ref{tour-path} vividly depicts
the tour path in guided tour. Guided tour, along with other types of
tour, has been implemented in the \emph{tourr} package in R, available
on the Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}.

@@

{[}doubt if I need the following. xxx - maybe an algorithmatic version
of guide tour would be helpful for non-expert to understand the
method?{]}

\hypertarget{optimisation-in-projection-pursuit}{%
\subsection{Optimisation in projection
pursuit}\label{optimisation-in-projection-pursuit}}

{[}thinking if the following three should be presented as algorithms or
plain description is fine. xxx{]}

Now we formulate the problem in mathmetical notation. Projection pursuit
finds the projection basis
\(\mathbf{A}^J = [\mathbf{a}_1, \cdots, \mathbf{a}_d]\), where
\(\mathbf{a}_i\) is a \(p \times 1\) vector, satisfying the following
optimisation problem:

@@

via an iterative search:
\(\{\mathbf{A}^1, \mathbf{A}^2, \cdots \cdots \mathbf{A}^J\}\). Guided
tour interpolates geodesically between each basis found by projection
pursuit \(A^j\) to form a tour path:
\[\{ \mathbf{A}^0, \mathbf{A}^1, {}_{1}\mathbf{A}^1, {}_{2}\mathbf{A}^1,\cdots, {}_{k_1}\mathbf{A}^1, A^2, {}_{1}\mathbf{A}^2, {}_{2}\mathbf{A}^2, \cdots {}_{k_2}\mathbf{A}^2, \cdots, \mathbf{A}^{J-1}, {}_{1}\mathbf{A}^{J-1}, \cdots {}_{k_m}\mathbf{A}^{J-1} ,\mathbf{A}^J \}\]

Where \(\mathbf{A}^0\) is the starting basis and \(\mathbf{A}^J\) is the
final basis that has the largest index value in the searching space.
Here we use left subscript to denote the index of the interpolation
bases in each iteration i.e. \({}_{2}\mathbf{A}^1\) means the second
interpolation basis in the first iteration. The right subcript is
reserved for the index of searching points in the optimisation
algorithm.

There are three existing methods for optimisating PPI function and we
review them below. \citet{posse1995projection} proposed a stochastic
direct search method, a random search algorithm. Given a current basis
\(\mathbf{A}^{j}\), a candidate basis \(\mathbf{A}^{j +1}_l\) is sampled
in the neighbourhood defined by the radius of the p-dimensional sphere,
\(\alpha\) of the current basis \(\mathbf{A}^{j}\) by
\[\mathbf{A}^{j+1}_{l} = \mathbf{A}^{j} + \alpha \mathbf{A}_{rand}\]. If
the candidate basis has a higher index value than the current basis, it
is outputed as the target basis \(\mathbf{A}^{j + 1}\) along with other
metadata and the current iteration stops. If no basis is found to have
higher index value after the maximum number of tries \(l_{\max}\), the
algorithm stops with nothing outputted. \(c\) is the halfing parameter
and when \(c > 30\), the nieghbourhood parameter \(\alpha\) in the next
iteration will be reduced by half. The algorithm for a random search is
summarised in Algorithm \ref{random-search}.

\newpage
\begin{algorithm}[H]
\SetAlgoLined
  initialisation: $c = 0$, $l = 1$ $l_{\max}$, $\mathbf{A}^{j}$\;
  \While{$l < l_{\max}$}{
    $\mathbf{A}^{j+1}_{l} = \mathbf{A}^{j} + \alpha \mathbf{A}_{rand}$\;
    $I^{j+1}_l = f(\mathbf{A}^{j+1}_l)$\;
    \eIf{$I^{j+1}_l > I^{j}$}{
      $\mathbf{A}^{j+1} = \mathbf{A}^{j+1}_{l}$\;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search: my first algorithm}
  \label{random-search}
\end{algorithm}

\citet{cook1995grand} explained the use of a gradient ascent
optimisation with the assumption that the index function is continuous
and differentiable. Since some indices could be non-differentiable, the
computation of derivative is replaced by a pseudo-derivative of
evaluating five randomly generated directions in a tiny nearby
neighbourhood. Taking a step on the straight derivative direction has
been modified to maximise the projection pursuit index along the
geodesic direction.

\newpage
\begin{algorithm}[H]
\SetAlgoLined
  initialisation: $l_{\max}$, $\alpha$, $l_{\max}$,$\mathbf{A}^j$\;
  \While{$l < l_{\max}$}{
    Generate $\{A^{(j_1)}: A^{(1)}, \cdots A^{(5)}\}$\;
    Find $A^{J_1}_l$ where $J_1 \in j_1$ such that $\{A^{(J_1)}: \arg \max I(A^{(j_1)})\}$\;
    Denote $F_a = A^j$ and $F_z = A^{(J_1)}$\;
    Conduct singular value decomposition (SVD) for $F_a$ and $F_z$ such that $F_{a}^{T}F_z = V_a\Lambda V_z^T$\;
    Compute $G_a = F_aV_a$, $G_z = F_zV_z$ and $\tau = \arccos(\Lambda)$\;
    ...\;
    Find $A^{(J_2)}$ satisfying $A^{J_2}$ on the interpolation path such that $\{A^{(J_2)}: \arg \max I(A^{(j_2)})\}$\;
    Compute $I^{(J_2)} = f(A^{(J_2)})$, $p_{diff} = (I^{(J_2)} - I^{j})/I^j$\;
      \If{$p_{diff} > 0.001$}{
        $A^{j+1} = A^{(J_2)}$\;
        $I^{j+1} = I^{(J_2)}$\;
        \KwOut{$A^{j+1}$, $I^{j+1}$, $i$ and relevant metadata}
      }
    $i = i + 1$\;
  }
  \caption{searchgeodesic: this part has problem!}
  \label{search_geodesic}
\end{algorithm}

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\)
is defined as \(T(l) = \frac{T_0}{log(l + 1)}\). The simulated annealing
algorithm works as follows. Given a neighbourhood parameter \(\alpha\)
and a randomly generated orthonormal basis \(A_{rand}\), a candidate
basis is constructed as
\(A^{j+1}_l = (1 - \alpha)A^{j} + \alpha A_{rand}\). If the index value
of the candidate basis is larger than the one of the current basis, the
candidate basis becomes the target basis. If it is smaller, the
candidate is accepted with probability
\(\text{prob} = \min\left\{\exp\left[\frac{I^j - I^{j+1}_l}{T(l)}\right],1\right\}\)
where \(I^j\) and \(I^{j+1}_l\) are the index value of the current and
candidate basis respectively. The algorithm can be summarised as in
Algorithm \ref{simulated_annealing}.

\newpage
\begin{algorithm}[H]
\SetAlgoLined
  initialisation: $T_0$, $\alpha$, $l_{\max}$,$\mathbf{A}^j$\;
  \While{$l < l_{\max}$}{
    $\mathbf{A}^{j+1}_{l} = (1 - \alpha)\mathbf{A}^{j} + \alpha \mathbf{A}_{rand}$ s.t. $\lVert\mathbf{a_i} \rVert = 1$\;
    $I^{j+1}_l = f(\mathbf{A}^{j+1}_{l})$\;
    $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I^{j+1}_l > I^{j}$}{
        $\mathbf{A}^{j+1} = \mathbf{A}^{j+1}_{l}$\;
        $I^{j+1} = I^{j+1}_l$\;
        \KwOut{$\mathbf{A}^{j+1}$, $I^{j+1}$,$l$ and relevant metadata}
      }{
        $\text{prob} = \min\left\{\exp\left[\frac{I^j - I^{j+1}_l}{T(l)}\right],1\right\}$\;
        Draw $\text{rand} \sim \text{Unif(0, 1)}$\;
        \If{$\text{prob} > \text{rand}$}{
          \KwOut{$\mathbf{A}^{j+1}$, $I^{j+1}$, $l$ and relevant metadata}
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Ggplot requires a data frame that contains all the variables to plot and
a \emph{global object} is constructed as the data frame supplied to the
visual diagnostic plots to better suit the characters of iterative
optimisation algorithms. Given an optimisation algorithm, two primary
variables of interest are the \emph{decision variable} and the
\emph{value of the objective function}. To align with the notation used
previously for projection pursuit, we denote the decision variable to be
\(\mathbf{A}\), which are matrices in projection pursuit algorithms and
value of the objective function to be \(I\), the index function.

\emph{Iterators} indexes the data collected and has a time series
feature that prescribes the its natural ordering. Given the complexity
of the algorithm, there could be multiple iterators. The simpliest
iterator: \emph{index per observation}, is a unique identifier for each
observation in the data. For each level of iteration, we design two
indices: \emph{index per iteration} is fixed for each observations in
one iteration and has an increment of one once a new iteration starts.
\emph{index within iteration} has an increment of one for each
observation in an iteartion and starts over from one once a new
iteration starts. A sketch of the difference between these three
iterators is provided in Figure \ref{iterators}. Depend on the number of
iteration level (\(L\)), the total number of iterators will differ and
can be calculated as \(1 + 2L\).

@@

There could exist other parameters that are also of our interest but
can't be classified as one of the three categories. They are defined
under the fourth category: \emph{other parameter of interest: \(O\)}. In
projection pursuit, three other parameters of our interest includes
\texttt{method}, \texttt{alpha} and \texttt{info}. \texttt{method}
identifies the name of the searching method used and we are interested
in comparing the performance between different algorithms under direct
search. The neighbourhood parameter \texttt{alpha} controls the size of
the sampling space and we are interested to understand how the searching
space shrinks as the algorithm progresses. \texttt{info} labels
different stages in the searching process. A sketch of the global object
for projection pursuit guided tour is presented in Figure
\ref{fig:glb-obj}.

@@

Now we formulate the idea of global object using mathmetical notation.
Let \[\mathbf{A}_{j,l,k}\] denotes the projection matrix in iteration
\(j\). If it is searching basis produced in the optimisation, then it
has a subscript \(l\). If it is an interpolation basis, it has a \(k\)
subscript. For example \(\mathbf{A}_{1,2,/}\) means the second searching
basis in the first iteration, \(\mathbf{A}_{1, /, 2}\) means the second
interpolation basis that on the tour path from the current basis
\(\mathbf{A}^0\) to the target basis found in iteration 1:
\(\mathbf{A}^1\). The target basis found in each iteration,
\(\mathbf{A}_{j, l_j, /}\) can also be simplified using a superscript as
\(\mathbf{A}^j\).

Let \(\mathbf{A}^0\) denotes the starting basis. Superscript \(j\) is
used to denote the number of iteration (index per iteration). Candidate
basis are denoted using subscript on the right and interpolation basis
are denoted using subscript on the left. For example,
\(\mathbf{A}^1_{2}\) means the second candidate basis in the first
iteration and \({}_{2}\mathbf{A}^1\) means the second interpolation
basis that forms the tour path from the current basis \(\mathbf{A}^0\)
to the target basis \(\mathbf{A}^1\). In iteration two, \(\mathbf{A}^1\)
becomes the current basis and projection pursuit searches through
\(\mathbf{A}^2_{1}, \mathbf{A}^2_{2}, \cdots, \mathbf{A}^2_{l_2}\) to
find the target basis \(\mathbf{A}^2\). The tour path from
\(\mathbf{A}^1\) to \(\mathbf{A}^2\) is constructed using the
interpolation bases
\({}_{1}\mathbf{A}^2, {}_{2}\mathbf{A}^2, \cdots, {}_{k_2}\mathbf{A}^2\).
All the bases together can be arranged in a format that each coloumn
represents an iteration as in Equation \ref{eq:matrixA}.

\begin{equation}
\left[
\begin{array}{ccccccc}
\ast & \mathbf{A}_{1,1,\ast} & \mathbf{A}_{2, l_1 +1,\ast} & \cdots & \mathbf{A}_{J,\sum_{i = 1}^{J-1}l_i +1,\ast} & \mathbf{A}_{J+1,\sum_{i = 1}^{J}l_i +1 ,\ast} \\
\ast & \mathbf{A}_{1,2,\ast} & \mathbf{A}_{2,l_1 +2,\ast} & \cdots & \mathbf{A}_{J,\sum_{i = 1}^{J-1}l_i +2,\ast} & \mathbf{A}_{J+1,\sum_{i = 1}^{J}l_i +2, \ast} \\
\vdots & \vdots & \vdots & \ddots & \vdots &\vdots   \\
\ast & \mathbf{A}_{1, l, \ast} & \mathbf{A}_{2, l_1 + l, \ast} & \cdots & \mathbf{A}_{J+1, \sum_{i = 1}^{J-1}l_i + l, \ast}  & \mathbf{A}_{J + 1, \sum_{i = 1}^{J}l_i + l, \ast}\\
\vdots & \vdots & \vdots & \ddots & \vdots &\vdots   \\
\ast & \mathbf{A}_{1, l_1-1, \ast} & \mathbf{A}_{2, l_1 + l_2-1, \ast} & \cdots & \mathbf{A}_{J, \sum_{i = 1}^{J-1}l_i -1, \ast} & \mathbf{A}_{J+1, \sum_{i = 1}^{J+1} l_i-1, \ast}\\
\mathbf{A}_{0,\ast,\ast} & \mathbf{A}_{1, l_1, \ast} & \mathbf{A}_{2, l_1 + l_2, \ast} & \cdots & \mathbf{A}_{J, \sum_{i = 1}^{J}l_i , \ast} & \ast\\
\hline
\ast &\mathbf{A}_{1, \ast, 1} & \mathbf{A}_{2, \ast, k_1 + 1} & \cdots &\mathbf{A}_{J, \ast,\sum_{i = 1}^{J-1}k_i + 1}  & \ast\\
\ast &\mathbf{A}_{1, \ast, 2} & \mathbf{A}_{2, \ast, k_1 + 2} & \cdots &\mathbf{A}_{J, \ast, \sum_{i = 1}^{J-1}k_i +2}  & \ast \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\ast &\mathbf{A}_{1, \ast, k} & \mathbf{A}_{2, \ast, k_1 + k} & \cdots &\mathbf{A}_{J, \ast, \sum_{i = 1}^{J-1}k_i +k}  & \ast \\
\vdots & \vdots & \vdots & \ddots & \vdots   & \vdots \\
\ast &\mathbf{A}_{1, \ast, k_1} & \mathbf{A}_{2, \ast, k_1 + k_2} & \cdots &\mathbf{A}_{J, \ast,\sum_{i = 1}^{J}k_i} & \ast \\
\end{array}
\right]
= 
\left[
\begin{array}{c}
\vphantom{A^J_1}\text{candidate basis} \\
\vphantom{A^J_2}\text{candidate basis} \\
\vphantom{\vdots} \vdots \\
\vphantom{A^J_2}\text{candidate basis} \\
\vphantom{\vdots} \vdots \\
\vphantom{A^J_2}\text{candidate basis} \\
\vphantom{A^J}\text{current/target basis} \\
\hline
\vphantom{\text{NULL}}\text{interpolation basis} \\
\vphantom{\text{NULL}}\text{interpolation basis} \\
\vphantom{\text{NULL}}\vdots \\
\vphantom{\text{NULL}}\text{interpolation basis} \\
\vphantom{\vdots} \vdots \\
\vphantom{\text{NULL}}\text{interpolation basis} \\
\end{array}
\right]
\label{eq:matrixA}
\end{equation}

When aggregated with information of index value and other interesting
parameter, it is easier to write the basis as a column vector denoting
by a single subscript by their natural ordering i.e.~all the bases in
iteration 1 along with the starting basis as shown in Equation
\ref{eq:vectorA}, can be denoted as
\(\mathbf{A}_0, \mathbf{A}_1, \cdots, \mathbf{A}_n\) where
\(n = 1 + l_1 + k_1\). The full data structure can thus be presneted as
in Equation \ref{eq:data-structure}.

\begin{equation}
\begin{array}{cccccccccccc}
\bigl[
\mathbf{A}_{0, \ast, \ast} & 
\mathbf{A}_{1, 1, \ast} & 
\mathbf{A}_{1,2,\ast} & 
\cdots & 
\mathbf{A}_{1,l_1, \ast}   &
\mathbf{A}_{1, \ast, 1} & 
\mathbf{A}_{1, \ast, 2} & 
\cdots &
\mathbf{A}_{1, \ast, k_1} &
\bigr]
\end{array} 
\label{eq:vectorA}
\end{equation}

\begin{equation}
\left[
\begin{array}{c|c|cccc|ccc}
\mathbf{A} & I & j &  l  & k & n & S_{1}& \cdots & S_{p}\\
\hline
\mathbf{A}_0 & I_0 & 0 & \ast & \ast & 1 & \ast & \cdots & \ast\\
\hline
\mathbf{A}_1 & I_1 & 1 & 1 & \ast& 2 & S_{1, 1} & \cdots & S_{1, p}\\
\mathbf{A}_2 & I_2 & 1 & 2 &\ast & 3 & S_{2, 1} & \cdots & S_{2, p}\\
\vdots &\vdots & \vdots &\vdots & \vdots &\vdots &\vdots &\cdots &\vdots\\
\vdots & \vdots & 1 & l_1 & \ast & \vdots & \vdots & \cdots & \vdots\\
\hline
\vdots & \vdots & 1 & \ast & 1& \vdots &\vdots & \cdots & \vdots\\
\vdots &\vdots& 1 & \ast & 2& \vdots&\vdots & \cdots & \vdots\\
\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\vdots &\cdots &\vdots \\
\vdots &\vdots & 1 & \ast & k_1&\vdots &\vdots & \cdots & \vdots\\
\hline
\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\vdots &\cdots &\vdots \\
\mathbf{A}_n & I_n & J &  l_J & \ast & n & S_{n, 1}& \cdots & S_{n, p}\\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

where
\(n = 1 + \sum_{i = 1}^{J+1} l_i -1 + \sum_{i = 1}^{J} k_i= \sum_{i = 1}^{J+1} l_i + \sum_{i = 1}^{J} k_i\)

>>>>

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. Let \(\mathbf{X}_{n \times p}\) be a
data matrix, an n-d projection can be seen as a linear transformation
\(T: \mathbb{R}^p \mapsto \mathbb{R}^d\) defined by
\(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis. Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) be an index function
that maps the projection basis \(\mathbf{A}\) onto a scalar value \(I\),
this function is commonly known as the projection pursuit index (PPI)
function, or the index function and can be used to measure the
``interestingness'' of the projection. A number of indice functions have
been proposed in the literature including Legendre index
\citep{friedman1974projection}, hermite index
\citep{hall1989polynomial}, natural hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}.

@@

Projection pursuit is usually used in conjunction with a tour method
called \emph{guided tour}. Tour explores the multivariate data
\emph{interactively} via playing a series of projections, that form a
\emph{tour path} and guided tour uses the path that is geodesically the
shortest. Details of the mathematical construction of a tour path can be
found in \citet{buja2005computational}. Guided tour, along with other
types of tour, has been implemented in the \emph{tourr} package in R,
available on the Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}.

For a projection basis, let the three-subscript notation
\[\mathbf{A}_{jlk}\] represent the projection matrix in iteration \(j\)
with either a searching index \(l\) or an interpolation index \(k\). The
placeholder \(*\) will be the substitue for the non-applicable index. To
give two examples, \(\mathbf{A}_{12*}\) denotes the second searching
basis in the first iteration; \(\mathbf{A}_{1*2}\) denotes the second
interpolation basis in the first iteration.

If we are being pedantic, all the target bases, except the starting
basis, should have a searching index since it is the last candidate
basis that successfully allows the optimisation algorithm to end and
output a new basis for interpolation. If we denotes the length of each
search (number of basis searched in each iteration) as
\(l_1, l_2, \cdots, l_J\) and the length of each interpolation is
\(k_1, k_2, \cdots, k_J\) , then the searching index for target basis in
each iteration will become
\(l_1, l_1 + l_2, \cdots, l_1 +l_2+\cdots l_J\). To avoid this complex
notation involving opteration in the subscript, we simplify it using a
superscript as
\[\mathbf{A}^{0}, \mathbf{A}^{1}, \mathbf{A}^{2}, \cdots, \mathbf{A}^{J}\]
where \(\mathbf{A}^{0}\) is the starting basis. Notice that in the case
when the complex notation can't be avoided, we shall separate each index
using a comma and write the index in brackets if necessary. Using the
notation above, all the bases on the interpolation path (including
target and interpolation bases) can be written as in Equation
\ref{eq:interp-path}.

\begin{equation}
\{ 
\mathbf{A}^0, 
\mathbf{A}^1, 
\mathbf{A}_{1 \ast 1}, 
\mathbf{A}_{1 \ast 2},
\cdots, 
\mathbf{A}^2, 
\mathbf{A}_{2 \ast 1}, 
\mathbf{A}_{2 \ast 2}, 
\cdots, 
\mathbf{A}^J,
\mathbf{A}_{J \ast 1}, 
\mathbf{A}_{J \ast 2}, 
\cdots, 
\mathbf{A}_{J \ast K},
\}
\label{eq:interp-path}
\end{equation}

where \(K\) denotes the last interpolating index. A visual
representation of the interpolation path ( for iteration one) modified
from \citet{buja2005computational} is shown in Figure \ref{tour-path}.
If we use a dot to represent a plane in the searching space, the
searching points along with the current and target basis in iteration
one can sketched in Figure \ref{searching-points}. Combine the two
above, we can write all the bases in the first iteration (including the
starting basis) in its natural ordering as in Equation \ref{eq:vectorA}.

\begin{equation}
\{
\mathbf{A}^{0}, 
\mathbf{A}_{1 1\ast}, 
\mathbf{A}_{1 2\ast}, 
\cdots, 
\mathbf{A}^{1},
\mathbf{A}_{1\ast 1}, 
\mathbf{A}_{1\ast 2},  
\cdots, 
\mathbf{A}_{1\ast k_1}
\}
\label{eq:vectorA}
\end{equation}

@@

\begin{figure}
\includegraphics[width=1\linewidth,height=0.6\textheight]{/Users/hzha400/Documents/3.PhD/paper-tour-vis/figures/tour_path_increasing} \caption{\label{searching-points}An illustration of the tour path}\label{fig:searching-points}
\end{figure}

\hypertarget{optimisation-problem-and-existing-algorithms}{%
\subsection{Optimisation problem and existing
algorithms}\label{optimisation-problem-and-existing-algorithms}}

Now we begin to formulate the optimisation problem. Given a randomly
generated starting basis \(\mathbf{A}^0\), projection pursuit finds the
final projection basis
\(\mathbf{A}^J = [\mathbf{a}_1, \cdots, \mathbf{a}_d]\), where
\(\mathbf{a}_i\) is a \(p \times 1\) vector, satisfies the following
optimisation problem:

@@

via an iterative search of target basis:
\(\{\mathbf{A}^0, \mathbf{A}^1, \mathbf{A}^2, \cdots \cdots \mathbf{A}^J\}\).

There are three existing methods for optimisating PPI function and we
review them below.

\citet{posse1995projection} proposed a stochastic direct search method,
a random search algorithm. Given a current basis \(\mathbf{A}^{j-1}\), a
candidate basis \(\mathbf{A}_{jl*}\) is sampled in the neighbourhood
defined by the radius of the p-dimensional sphere, \(\alpha\) of the
current basis \(\mathbf{A}^{j}\) by
\(\mathbf{A}_{jl*} = \mathbf{A}^{j-1} + \alpha \mathbf{A}_{rand}\). If
the candidate basis has a higher index value than the current basis, it
is outputed as the target basis \(\mathbf{A}^{j}\) along with other
metadata and the current iteration stops. If no basis is found to have
higher index value after the maximum number of tries \(l_{\max}\), the
algorithm stops with nothing outputted. \(c\) is the halfing parameter
and when \(c > 30\), the nieghbourhood parameter \(\alpha\) in the next
iteration will be reduced by half. The algorithm for a random search is
summarised in Algorithm \ref{random-search}.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{jl*} = \mathbf{A}^{j-1} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{j, l,*}$ is orthonormal\;
    Compute $I_{jl*}  = f(\mathbf{A}_{jl*})$\;
    \eIf{$I_{jl*} > I^{j-1}$}{
      $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

\citet{cook1995grand} explained the use of a gradient ascent
optimisation with the assumption that the index function is continuous
and differentiable. Since some indices could be non-differentiable, the
computation of derivative is replaced by a pseudo-derivative of
evaluating five randomly generated directions in a tiny nearby
neighbourhood. Taking a step on the straight derivative direction has
been modified to maximise the projection pursuit index along the
geodesic direction. See Algorithm \ref{search-geodesic} for a
summarisation of the steps.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate ten bases in five random directions: $\mathbf{A}_{jl*}: \mathbf{A}_{j,(l + 9), *}$ within a small neighbourhood $\text{dist}$\;
    Find the direction with the largest index value: $\mathbf{A}_{j, l_{d}, *}$\;
    Construct the geodesic from $\mathbf{A}^{j-1}$ to $\mathbf{A}_{j, l_{d}, *}$\;
    Find $\mathbf{A}_{j,l_g, *}$ on the geodesic that has the largest index value \;
    Compute $I_{j,l_g, *} = f(\mathbf{A}_{j,l_g, *})$, $p_{\text{diff}} = (I_{j,l_g, *} - I^{j-1})/I_{j,l_g, *}$\;
      \If{$p_{\text{diff}} > 0.001$}{
        $A^{j} = A_{j,l_g, *}$, $I^{j} = I_{j,l_g, *}$\;
      }
    $l = l + 1$\;
  }
  \caption{search geodesic}
  \label{search-geodesic}
\end{algorithm}

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\)
is defined as \(T(l) = \frac{T_0}{log(l + 1)}\). The simulated annealing
algorithm works as follows. Given a neighbourhood parameter \(\alpha\)
and a randomly generated orthonormal basis \(A_{rand}\), a candidate
basis is constructed as
\(\mathbf{A}_{jl*} = (1 - \alpha)\mathbf{A}^{j} + \alpha \mathbf{A}_{rand}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis becomes the target basis. If it is
smaller, the candidate is accepted with probability
\(P= \min\left\{\exp\left[\frac{I^{j-1} - I_{jl*}}{T(l)}\right],1\right\}\)
where \(I^{j-1}\) and \(I_{jl*}\) are the index value of the current and
candidate basis respectively. The algorithm can be summarised as in
Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{jl*} = (1 - \alpha)\mathbf{A}^{j-1} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{jl*}$ is orthonormal \;
    Commpute $I_{jl*} = f(\mathbf{A}_{jl*})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{jl*} > I^{j-1}$}{
        $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I^{j-1} - I_{jl*}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
          $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Ggplot requires a data frame that contains all the variables to plot and
a \emph{global object} is constructed as the data frame supplied to the
visual diagnostic plots to better suit the characters of iterative
optimisation algorithms. Given an optimisation algorithm, two primary
variables of interest are the \emph{decision variable: \(\mathbf{A}\)}
and the \emph{value of the objective function: \(I\)}.

\emph{Iterators} indexes the data collected and has a time series
feature that prescribes the its natural ordering of the decision varible
in the searching. The simpliest iterator, \emph{index per observation:
\(t\)}, is a unique identifier for each observation in the data. For
each level of iteration, we design two types of indices: \emph{index per
iteration} is fixed for each observations in one iteration and has an
increment of one once a new iteration starts. \emph{index within
iteration} has an increment of one for each observation in an iteartion
and starts over from one once a new iteration starts. A sketch of the
difference between these three iterators is provided in Figure
\ref{iterators}.

@@

There could exist other parameters that are also of our interest but
can't be classified as one of the three categories. They are defined
under the fourth category: \emph{other parameter of interest: \(S\)}. In
projection pursuit, three other parameters of our interest includes
\texttt{method}, \texttt{alpha} and \texttt{info}. \texttt{method}
identifies the name of the searching method used and we are interested
in comparing the performance between different algorithms under direct
search. The neighbourhood parameter \texttt{alpha} controls the size of
the sampling space and we are interested to understand how the searching
space shrinks as the algorithm progresses. \texttt{info} labels
different stages in the searching process. A sketch of the global object
for projection pursuit guided tour is presented in Figure
\ref{fig:glb-obj}.

@@

To write down the notation for the data structure, it is easier to write
the bases as a column vector denoting by a single subscript by their
natural ordering:

\begin{equation}
\{
\mathbf{A}_0, 
\mathbf{A}_1, 
\mathbf{A}_2, 
\cdots,
\mathbf{A}_T
\label{vector-basis}
\}
\end{equation}

where \(T\) is the total number of plane serached by the optimisation
algorithm including all the target planes, all the interpolating planes
and all the candidate planes but NOT the starting plane. The full data
structure can thus be shown as in Equation \ref{eq:data-structure}.

\begin{equation}
\left[
\begin{array}{c|cc|ccc|ccc}
t & \mathbf{A} & I & j &  l  & k & S_{1}& \cdots & S_{p}\\
\hline
1 & \mathbf{A}_0 & I_0 & 0 & \ast & \ast & \ast & \cdots & \ast\\
\hline
2 & \mathbf{A}_1 & I_1 & 1 & 1 & \ast & S_{1, 1} & \cdots & S_{1, p}\\
3 & \mathbf{A}_2 & I_2 & 1 & 2 & \ast & S_{2, 1} & \cdots & S_{2, p}\\
\vdots & \vdots &\vdots & \vdots &\vdots & \vdots &\vdots &\cdots &\vdots\\
\vdots & \vdots & \vdots & 1 & l_1 & \ast & \vdots &  \cdots & \vdots\\
\hline
\vdots &\vdots & \vdots & 1 & \ast & 1& \vdots &  \cdots & \vdots\\
\vdots &\vdots &\vdots& 1 & \ast & 2& \vdots& \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
\vdots &\vdots &\vdots & 1 & \ast & k_1&\vdots & \cdots & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
T+1 & \mathbf{A}_T & I_T & J &  l_J & \ast & S_{T, 1}& \cdots & S_{T, p}\\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}
